---
title: "Assignment 2: Coding Basics"
author: "Sam Vanasse"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 
```{r}
hundred.seq <- seq(1, 100, 4)
hundred.seq
```

2. Compute the mean and median of this sequence. 
```{r}
mean(hundred.seq)
median(hundred.seq)
```


3. Ask R to determine whether the mean is greater than the median. 
```{r}
mean(hundred.seq) < median(hundred.seq)
mean(hundred.seq) > median(hundred.seq)
mean(hundred.seq) != median(hundred.seq)
mean(hundred.seq) == median(hundred.seq)
```


4. Insert comments in your code to describe what you are doing.

```{r}
#1. Generating a sequence from 1 to 100, increasing the values by 4 and naming the sequence hundred.seq.

#2. Generating the mean and median of hundred.seq

#3. Demonstration that the mean and the medians are equal by showing the median is not greater than than the mean, the mean is not greater than the median, the two are not different, and finally that the two are equal. 

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
vector.names <- c("Carl", "William", "Matt", "Henry") #Character
vector.test.scores <- c(90,60,40,80) #Numeric
vector.pass.fail <- c(TRUE,TRUE,FALSE,TRUE) #Logical

student.dataframe <- cbind(vector.names,vector.test.scores,vector.pass.fail)
student.dataframe <- data.frame("Name"=vector.names,"Score"=vector.test.scores,"Passed"=vector.pass.fail)
student.dataframe

class(vector.names)
class(vector.test.scores)
class(vector.pass.fail)
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: A matrix requires equal objects in the rows and columns whereas a dataframe like above can have more rows than columns or vice-versa. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#Option 1
ifelse.test.score <- function(x){
  if(x<=49){"FAIL"}
  else {"PASS"}
}

score1 <- ifelse.test.score(90); score1
score2 <- ifelse.test.score(60); score2
score3 <- ifelse.test.score(40); score3
score4 <- ifelse.test.score(80); score4

score.all <- ifelse.test.score(vector.test.scores); score.all #Intentionally running an error to prove only the 'ifelse' statement can run all observations in a sector


#Option 2
ifelse.test.score2 <- function(x){
  ifelse(x>50, "PASS", "FAIL")
}
ifelse.score1 <- ifelse.test.score2(90); ifelse.score1
ifelse.score2 <- ifelse.test.score2(60); ifelse.score2
ifelse.score3 <- ifelse.test.score2(40); ifelse.score3
ifelse.score4 <- ifelse.test.score2(80); ifelse.score4

ifelse.score.all <- ifelse.test.score2(vector.test.scores); ifelse.score.all #Only this option works for this code

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: In order to use the if-else statement on all test scores in the vector, only the 'ifelse' statement will work. 
